 

### 2021-01-12 vertical-thermal-review.Rmd
Update vertical-thermal-review.Rmd
Minor edits throughout

If soil and stem respiration do not acclimate fully to rising temperatures, the greater amount of CO~2~ emitted at the forest floor (i.e. from soils and decaying organic matter and leaf litter) could increase the vertical CO~2~ gradient in the future, but this would still only be a short-term effect in the early morning, as the CO~2~ would quickly diffuse through the canopy. Furthermore, increases in mortality of big trees (e.g. Senf et al. 2018) would increase canopy roughness and prevent build-up of CO~2~ concentration gradients.
(Elsa: see these papers on the long-term Harvard Forest soil warming experiments. CO2 emissions increased, then decreased, then increased again. It's thought that the changes in soil CO2 emissions over time were linked to turnover in microbial community composition.
- DeAngelis, K.M., Pold, G., Topçuoğlu, B.D., van Diepen, L.T., Varney, R.M., Blanchard, J.L., Melillo, J. and Frey, S.D., 2015. Long-term forest soil warming alters microbial communities in temperate forest soils. Frontiers in microbiology, 6, p.104.
- Melillo, J.M., Steudler, P.A., Aber, J.D., Newkirk, K., Lux, H., Bowles, F.P., Catricala, C., Magill, A., Ahrens, T. and Morrisseau, S., 2002. Soil warming and carbon-cycle feedbacks to the climate system. Science, 298(5601), pp.2173-2176.
- Melillo, J.M., Frey, S.D., DeAngelis, K.M., Werner, W.J., Bernard, M.J., Bowles, F.P., Pold, G., Knorr, M.A. and Grandy, A.S., 2017. Long-term pattern and magnitude of soil carbon feedback to the climate system in a warming world. Science, 358(6359), pp.101-105.)

@@

Leaf-scale properties and processes aggregate non-linearly  in vertical and horizontal space, influencing stand-level ecosystem processes, and in turn biosphere-atmosphere exchanges of water, carbon, energy, and trace gases. As described above, variation in temperature across vertical gradients can interact with local light conditions and other vertically varying meteorological conditions to affect species composition (e.g., see Nakamura et al 2017), leaf traits, and photosynthesis . Linked to this local biophysical variation is the partitioning of evapotranspiration (ET) into soil evaporation, canopy evaporation, and transpiration, which occur across varying spatiotemporal scales depending on light availability, water stress, stomatal physiology, and turbulent transfer (Lawrence et al. 2007, Staudt et al. 2011). Vertical microclimate gradients interact with vertical variation in foliage density, influencing not only the location of evapotranspiration in the canopy, but also differences in rates and net gross primary productivity (GPP, Banerjee & Linn 2018). It has been demonstrated that net carbon assimilation and transpiration can vary substantially for forest canopies with the same LAI but different leaf area density distributions (e.g., Katul et al 2004 , Banerjee & Linn 2018). Further, although transpiration tends to be the dominant component of ET, partitioning is more complex in open canopies where soil evaporation can become more dominant (Baldocchi et al. 2004, Ma et al. 2020), likely affecting vertical thermal gradients. In other words, it matters whether foliage is dense or sparse, and whether it is concentrated higher in the canopy, lower in the canopy, or evenly distributed throughout.

@@

An increasing availability of airborne and spaceborne LiDAR remote sensing data offers a promising opportunity for mapping vertical thermal gradients and vegetation structure. [Drone based FLIR cams, tower thermal cams (Pau et al 2018), ECOSTRESS]. To evaluate large-scale ecosystem patterns, high resolution vertically resolved vegetation structure from LiDAR data can be analyzed in combination with collocated field measurements collected using towers (FluxNet REF), canopy cranes (Nakamura et al. 2017), or by tree climbing (REF – e.g. Asner papers, Shenkin et al – Unding Jami’s work). Existing research efforts have already revealed incredibly heterogeneity in vegetation structure, diversity, traits, and properties both within and across ecosystems (REFS – Stovall, Schneider, …others).

*Representing Vertical Gradients in Models* 
-	How do models currently integrate or not integrate these concepts and dynamics
-	What does that mean for about ability to predict/understand how forest ecosystems will respond in the future?

*Knowledge Gaps & Future Research*
-	What major gaps in knowledge and/or important questions emerge from this review (i.e. a lot of focus on droughts justifiably, but understanding heterogeneous impacts of temperature increases on forest ecosystems, particularly at vertically resolved scales is also important.)

E.g. see de Pury & Farquhar, 1997 : though likely an improvement over 'big-leaf' models, models that separate the canopy into sunlit and shaded portions (sun/shade models) may not be able to capture important within-canopy variation in terms of leaf dynamics (e.g. seasonal shifts in vertical leaf area distributions, Smith et al. 2019) and functions (e.g. thermal responses, as we present in this paper). Hence, multi-layered ecosystem models may be necessary for accurately projecting future forest function.

>>>>

If soil and stem respiration do not acclimate fully to rising temperatures, the greater amount of CO~2~ emitted at the forest floor (i.e. from soils and decaying organic matter and leaf litter) could increase the vertical CO~2~ gradient in the future, but this would still only be a short-term effect in the early morning, as the CO~2~ would quickly diffuse through the canopy. Furthermore, increases in mortality of big trees (e.g. Senf et al. 2018) would increase canopy roughness and prevent build-up of CO~2~ concentration gradients.
(Elsa: see these papers on the long-term Harvard Forest soil warming experiments. CO2 emissions increased, then decreased, then increased again. It's thought that the changes in soil CO2 emissions over time were linked to turnover in microbial community composition.
- DeAngelis, K.M., Pold, G., Topçuoğlu, B.D., van Diepen, L.T., Varney, R.M., Blanchard, J.L., Melillo, J. and Frey, S.D., 2015. Long-term forest soil warming alters microbial communities in temperate forest soils. Frontiers in microbiology, 6, p.104.
- Melillo, J.M., Steudler, P.A., Aber, J.D., Newkirk, K., Lux, H., Bowles, F.P., Catricala, C., Magill, A., Ahrens, T. and Morrisseau, S., 2002. Soil warming and carbon-cycle feedbacks to the climate system. Science, 298(5601), pp.2173-2176.
- Melillo, J.M., Frey, S.D., DeAngelis, K.M., Werner, W.J., Bernard, M.J., Bowles, F.P., Pold, G., Knorr, M.A. and Grandy, A.S., 2017. Long-term pattern and magnitude of soil carbon feedback to the climate system in a warming world. Science, 358(6359), pp.101-105.) 
- Nottingham, A.T., Meir, P., Velasquez, E. and Turner, B.L., 2020. Soil carbon loss by experimental warming in a tropical forest. Nature, 584(7820), pp.234-237.

@@

Leaf-scale properties and processes aggregate non-linearly  in vertical and horizontal space, influencing stand-level ecosystem processes, and in turn biosphere-atmosphere exchanges of water, carbon, energy, and trace gases. As described above, variation in temperature across vertical gradients can interact with local light conditions and other vertically varying meteorological conditions to affect species composition (e.g., see @nakamuraForestsTheirCanopies2017), leaf traits, and photosynthesis . Linked to this local biophysical variation is the partitioning of evapotranspiration (ET) into soil evaporation, canopy evaporation, and transpiration, which occur across varying spatiotemporal scales depending on light availability, water stress, stomatal physiology, and turbulent transfer (Lawrence et al. 2007, Staudt et al. 2011). Vertical microclimate gradients interact with vertical variation in foliage density, influencing not only the location of evapotranspiration in the canopy, but also differences in rates and net gross primary productivity (GPP, Banerjee & Linn 2018). It has been demonstrated that net carbon assimilation and transpiration can vary substantially for forest canopies with the same LAI but different leaf area density distributions (e.g., Katul et al 2004 , Banerjee & Linn 2018). Further, although transpiration tends to be the dominant component of ET, partitioning is more complex in open canopies where soil evaporation can become more dominant (Baldocchi et al. 2004, Ma et al. 2020), likely affecting vertical thermal gradients. In other words, it matters whether foliage is dense or sparse, and whether it is concentrated higher in the canopy, lower in the canopy, or evenly distributed throughout.

@@

An increasing availability of airborne and spaceborne LiDAR remote sensing data offers a promising opportunity for mapping vertical thermal gradients and vegetation structure. [Drone based FLIR cams, tower thermal cams (Pau et al 2018), ECOSTRESS]. To evaluate large-scale ecosystem patterns, high resolution vertically resolved vegetation structure from LiDAR data can be analyzed in combination with collocated field measurements collected using towers (FluxNet REF), canopy cranes (@nakamuraForestsTheirCanopies2017), or by tree climbing (REF – e.g. Asner papers, Shenkin et al – Unding Jami’s work). Existing research efforts have already revealed incredibly heterogeneity in vegetation structure, diversity, traits, and properties both within and across ecosystems (REFS – Stovall, Schneider, …others).

*Representing Vertical Gradients in Models* 
-	How do models currently integrate or not integrate these concepts and dynamics
-	What does that mean for about ability to predict/understand how forest ecosystems will respond in the future?
- E.g. see de Pury & Farquhar, 1997 : though likely an improvement over 'big-leaf' models, models that separate the canopy into sunlit and shaded portions (sun/shade models) may not be able to capture important within-canopy variation in terms of leaf dynamics (e.g. seasonal shifts in vertical leaf area distributions, Smith et al. 2019) and functions (e.g. thermal responses, as we present in this paper). Hence, multi-layered ecosystem models may be necessary for accurately projecting future forest function.

*Knowledge Gaps & Future Research*
-	What major gaps in knowledge and/or important questions emerge from this review (i.e. a lot of focus on droughts justifiably, but understanding heterogeneous impacts of temperature increases on forest ecosystems, particularly at vertically resolved scales is also important.)
- Not sure where this section should go
